Epigram 2\\ more design thoughts

### 2010-08-31 notes/Epigram2/Epigram2.tex
notes: stuff on interaction and where-clauses I tweaked the notation for sheds to be {. .} in order to allow {? ?} and {! !} for query and response. I've written a little about the uses of query and response in the coauthor transducer model.
I've added a bit about where-clauses, for helper-functions and
lemmas. None of this is terribly precise yet, and the notation
is very much up for grabs.

\documentclass{article}
\usepackage{palatino}
\usepackage{a4}

@@

Epigram documents show the declaration and partial solution of problems.
Problem solving is mediated as an interactive process. Every problem
in the source language can be encoded as a type in the evidence
language. Solving a problem in the source language amounts to finding
an inhabitant of the type which encodes it, by a process of
hierarchical refinement.

A \textbf{document} is a series of \textbf{developments}. A
development comprises a \textbf{declaration} introducing a problem,
and a \textbf{refinement} reducing the problem to a hopefully
empty collection of subproblems. Too much abstraction: example, please!
\newcommand{\capbox}[2]{\raisebox{0.12in}{\begin{array}[t]{|l|}\hline
      #1 \vspace*{-0.05in}\\
    \hfill\mbox{\scriptsize{#2}} \\ \hline\end{array}}}
\newcommand{\Ts}[1]{\texttt{#1}\;}
\newcommand{\co}{\;:\;}
\newcommand{\sco}{;\;\;}
\newcommand{\cm}{,\;}
\[
\capbox{
  \Ts{let}
    \capbox{
      \capbox{\Ts{x} \cm \Ts{y}\co \Ts{Nat}}{declaration}\\
      \Ts{--------------}\\
      \capbox{\Ts{x} \Ts{+} \Ts{y}}{template} \co \capbox{\Ts{Nat}}{type}
      }
      {declaration}\\ \\
  \capbox{
     \capbox{\Ts{x}\Ts{+}\Ts{y}}{problem}
       \capbox{\Ts{<=}\capbox{\Ts{induction}\Ts{x}}{eliminator}}{tactic}\\
     \capbox{
      \{\;\capbox{\capbox{\Ts{'zero}\Ts{+}\Ts{y}}{problem}
                  \capbox{\Ts{=} \capbox{\Ts{y}}{term}}{tactic}
                  \capbox{}{block}
          }{refinement}\\
      ;\;\capbox{\capbox{\Ts{'suc}\Ts{x'}\Ts{+}\Ts{y}}{problem}
                 \capbox{\Ts{=} \capbox{\Ts{'suc}
                   \capbox{\{!\;!\}}{shed}}{term}}{tactic}
                  \capbox{}{block}
         }{refinement}\\
      \}
      }{block}
    }
    {refinement}}
    {development}
\]

@@

The \textbf{shed}, \texttt{\{! !\}}, stands in place of a piece of program
(most syntactic categories, not just terms) to indicate that the programmer
has yet to choose something, thus causing elaboration to suspend. Note
that a tactic \texttt{= \{! !\}} does commit to return, where a
shed directly in a tactic place does not.

\end{document}

>>>>

\documentclass{article}
\usepackage{palatino}
\usepackage{a4}
\usepackage{alltt}

@@

Epigram documents show the declaration and partial solution of problems.
Problem solving is mediated as an interactive process. Every problem
in the source language can be encoded as a type in the evidence
language. Solving a problem in the source language amounts to finding
an inhabitant of the type which encodes it, by a process of
hierarchical refinement. This section is not supposed to be a definition:
it's an explorations.

A \textbf{document} is a series of \textbf{developments}. A
development comprises a \textbf{declaration} introducing a problem,
and a \textbf{refinement} reducing the problem to a hopefully
empty collection of subproblems. Too much abstraction: example, please!
\newcommand{\capbox}[2]{\raisebox{0.12in}{\begin{array}[t]{|l|}\hline
      #1 \vspace*{-0.05in}\\
    \hfill\mbox{\scriptsize{#2}} \\ \hline\end{array}}}
\newcommand{\Ts}[1]{\texttt{#1}\;}
\newcommand{\co}{\;:\;}
\newcommand{\sco}{;\;\;}
\newcommand{\cm}{,\;}
\[
\capbox{
  \capbox{
    \capbox{\Ts{x} \cm \Ts{y}\co \Ts{Nat}}{declaration}\\
    \Ts{--------------}\\
    \capbox{\Ts{x} \Ts{+} \Ts{y}}{template} \co \capbox{\Ts{Nat}}{type}
    }
    {declaration}\\ \\
  \capbox{
     \capbox{\Ts{x}\Ts{+}\Ts{y}}{problem}
       \capbox{\Ts{<=}\capbox{\Ts{induction}\Ts{x}}{eliminator}}{tactic}\\
     \capbox{
      \{\;\capbox{\capbox{\Ts{'zero}\Ts{+}\Ts{y}}{problem}
                  \capbox{\Ts{=} \capbox{\Ts{y}}{term}}{tactic}
                  \capbox{}{block}
          }{refinement}\\
      ;\;\capbox{\capbox{\Ts{'suc}\Ts{x'}\Ts{+}\Ts{y}}{problem}
                 \capbox{\Ts{=} \capbox{\Ts{'suc}
                   \capbox{\{.\;.\}}{shed}}{term}}{tactic}
                  \capbox{}{block}
         }{refinement}\\
      \}
      }{block}
    }
    {refinement}}
    {development}
\]

@@

The \textbf{shed}, \texttt{\{. .\}}, stands in place of a piece of program
(most syntactic categories, not just terms) to indicate that the programmer
has yet to choose something, thus causing elaboration to suspend. Note
that a tactic \texttt{= \{. .\}} does commit to return, where a
shed directly in a tactic place does not.

Now, in the coauthor transducer model, there is no trigger to fill in
a shed: one simply replaces the shed with code and reloads. Of course,
an editor could be customized to give this illusion, and the
patch-to-patch optimization would yield very similar results. What,
then, is the interaction?

The basic refinement interaction simply involves the system replacing
a null block with subproblems which are not accounted for.
Correspondingly,

\begin{alltt}
 x , y : Nat
-------------
 x + y : Nat

x + y <= induction x
\end{alltt}
%
gets fleshed out to
%
\begin{alltt}
let   x , y : Nat
     -------------
      x + y : Nat

x + y <= induction x
\{ 'zero + y    \{. .\}
; 'suc x' + y  \{. .\} 
\}
\end{alltt}

One might imagine going further, reporting coverage failure by inserting
subproblems into nonempty blocks. Such a circumstance might readily
arise if one added constructors to a datatype.

\subsection{Query and Response}

In addition to basic refinement functionality, we shall need to equip
programmers with the means to query the system and receive responses
in situ. Moreover, we shall need to equip the machine with the means
to insert remarks in a file. We need anchored comments. The machine
writes
\[
\Ts{\{!}\; \mathit{thing}\; \Ts{!!} \mathit{comment} \Ts{!\}}
\]
to indicate that the \(\mathit{comment}\) concerns the \(\mathit{thing}\).
Such a textual pattern is replaced by \(\mathit{thing}\) on reloading
(unless \(\mathit{thing}\) is blank, in which case the replacement is an
empty shed). Machine remarks are ephemeral responses, and should not be
put into the machine's mouth by the programmer. Meanwhile, an anchored query,
\[
\Ts{\{?}\; \mathit{thing}\; \Ts{??} \mathit{query} \Ts{?\}}
\]
means the same as \(\mathit{thing}\) (unless empty, in which case it means
the same thing as an empty shed) but attaches a query; on return, we can
expect a response in the form given above.

Your basic empty \texttt{\{? ?\}} means
``What goes here and what's in local scope?''. The basic use of
\(\Ts{\{!}\; \mathit{thing}\; \Ts{!!} \mathit{comment} \Ts{!\}}\)
is to report an error. A more amusing possibility would be to offer
a menu of choices for \(\mathit{thing}\).

It does not take a great leap of imagination to imagine how a fancier
interface might mediate point-and-click queries via this mechanism.
Again, one would need the patch-based transducer to support such a
rapid dialogue.

\subsection{Subproblems versus Side-Problems}

The hierarchical refinement of problems delivers the basic necessary
functionality, but it forces all sorts of auxiliary functions and lemmas
to be developed globally, in advance (spatially, if not temporally).
We might prefer to localize the development of helpers. I suggest that
we allow each refinement to carry a \texttt{where}-clause bearing a local
module of (partially) solved side-problems. Here's a simple example.

\begin{alltt}
 f : X -> Y; xs : List X
-------------------------
 map f xs : List Y

map f xs = hit xs where
\{  xs : List X
  -----------------
   hit xs : List X

  hit xs <= induction xs
  \{ hit 'nil = 'nil
  ; hit ('cons x xs') zs = 'cons (f x) (hit xs') \}\}
\end{alltt}

Scope-wise, the \texttt{where}-clause comes after the problem statement
but before tactic and block; textually, it can come before the tactic,
between tactic and block, or after block.

Note that on a good day with a following wind, the types of helper
functions can be inferred. One might hope to write
\begin{alltt}
 f : X -> Y; xs : List X
-------------------------
 map f xs : List Y

map f xs = \{? hit xs ?? helper ?\}
\end{alltt}
might deliver
\begin{alltt}
 f : X -> Y; xs : List X
-------------------------
 map f xs : List Y

map f xs = hit xs where
\{  xs : List X
  -----------------
   hit xs : List X

  hit xs \{! !\} \}
\end{alltt}

on the basis that the type \texttt{List X} is being pushed into
the helper application, comprising the function \texttt{hit} applied
(Miller-pattern-style) to distinct local variables (here just \texttt{xs})
whose types are known, thus determining \texttt{hit}'s declaration!
Where the helper function has need of implicit syntax, a little more
care will be needed to ensure control over what gets abstracted.

Propositional requirements arising in the course of programming might
also be discharged as side-problems in \texttt{where}-clauses. We do not
write proofs in terms: we rather expect proofs to be picked up from the
context, so side-problems allow us to seed the context appropriately.
Moreover, elaboration models type constraints as propositional equations,
so you can fix type-glitches (where provably equal types mismatch
definitionally) by adding the proofs to a \texttt{where}-clause.
\end{document}
